\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-12-11 analysis/paper/paper.tex

To conduct our review on current methods available in survival analysis,
a scraper was coded to examine all packages in the survival task view on
CRAN \citeyearpar{survTaskView}. There was a total of 259 packages, many
of which are not relevant to the problems casebase addresses. To filter
these packages, a scraper was used to search for competing risks,
cumulative incidence functions, non- proportional hazards and
penalization. This reduced our set of 259 packages down to 60. From
here, we manually examined each package to determine which ones are
comparable to casebase, with either a different approach to the same
problems or future features we would like to add to casebase.

These packages can belong to three main categories: non-proportional
hazards, regularization and competing risks. Besides these three main
categories, there are extra features of interest as well. Table XXX
summarizes the findings.

Many packages exist that implement semi-parametric cox models. Of those,
glmnet \citeyearpar{regpathcox}???, glmpath \citeyearpar{park_hastie},
penalized \citeyearpar{l1penal}, RiskRegression
\citeyearpar{gerds_blanche} introduce penalization in their model
fitting. CoxRidge \citeyearpar{perperoglou} , rstpm2
\citeyearpar{clements_liu} and survival \citeyearpar{survival-package}
include penalization, but they implement the extended cox model
permitting intuitive non proportional hazard modeling while the previous
four do not. In terms of penalization, the benefit of casebase comes
from its flexibility. Linear models have a wide tool set, considerably
larger than those made for cox regression (\emph{reference for number of
packages between their task views, regression is spread out across
Multivariate, SocialSciences, Robust among others. The focus of this
task view is on Optimization Infrastructure Packages , General Purpose
Continuous Solvers , Mathematical Programming Solvers , and Specific
Applications in Optimization , or Multi Objective Optimization}).
Casebase permits many regularization tools developed for linear
regression to be repurposed within a survival context.

Parametric models permit non-proportional hazards and are implemented in
a handful of packages. Of interest are CFC
\citeyearpar{sharabiani_mahani_2019}, flexsurv \citeyearpar{flexsurv} ,
SmoothHazard \citeyearpar{smoothHazard}, Rsptm2
\citeyearpar{clements_liu} and Survival \citeyearpar{survival-package}.
Smooth hazard is limited to weibull
distributions\citeyearpar{smoothHazard}, where as flexsurv and survival
contain implementations for users to supply any distribution they would
like \citeyearpar{survival-package}, \citeyearpar{flexsurv}. Flexsurv,
smooth hazard, mets and rstpm2 can use splines, permitting a flexible
fit over time \citeyearpar{flexsurv}, \citeyearpar{smoothHazard},
\citeyearpar{scheike2014estimating}, \citeyearpar{clements_liu} . Though
casebase is limited to 3 distributions at the moment, it also uses
splines allowing for distributions that may not be consistent over
survival time.

Of the methods mentioned so far, only CFC,flexsurv, mets and survival
contain implementations for competing risks
\citeyearpar{sharabiani_mahani_2019},
\citeyearpar{flexsurv},\citeyearpar{scheike2014estimating}
,\citeyearpar{survival-package} (\emph{how is it done for parametric
models?}). The differentiating factor between these packages and
casebase is that casebase handles competing risks through multiple
logistic regression. Smooth hazard, survival and Rstpm2 can handle
interval censoring, where as casebase currently can not. Interval
censoring is simple feature to implement and is a potential improvement
to add to casebase. (There's also multistate models, which I haven't
mentioned in this paper)

Of all the packages listed in the table, CFC , Flexsurv, mets,
RiskRegression, rstpm2 and survival have implementations for cumulative
incidence \citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche},
\citeyearpar{clements_liu}, \citeyearpar{survival-package}. Since
casebase is parametric, the baseline hazard is known and requires
integration to retrieve the cumulative incidence. This is similar to the
approaches of CFC and flexsurv \citeyearpar{sharabiani_mahani_2019},
\citeyearpar{flexsurv}, while the other packages use semi-parametric
approaches to retrieve the baseline after fitting
\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche}. rstpm2
and survival use both for their respective parametric and
semi-parametric model implementations \citeyearpar{clements_liu},
\citeyearpar{survival-package}.

\begin{table}[]

@@

This report was generated on 2019-12-11 12:28:29 using the following
computational environment and dependencies:

@@

#> ─ Session info ───────────────────────────────────────────────────────────────

@@

#>  tz       America/Montreal            

@@

#> ─ Packages ───────────────────────────────────────────────────────────────────

@@

#>  callr         3.4.0      2019-12-09 [1]

@@

#>  codetools     0.2-16     2018-12-24 [1]

@@

#>  DBI           1.0.0      2018-05-02 [1]
#>  dbplyr        1.4.2      2019-06-17 [1]

@@

#>  devtools      2.2.1      2019-09-24 [1]

@@

#>  ellipsis      0.3.0      2019-09-20 [1]

@@

#>  foreach       1.4.7      2019-07-27 [1]

@@

#>  glmnet      * 3.0-1      2019-11-15 [1]

@@

#>  haven         2.2.0      2019-11-08 [1]

@@

#>  htmltools     0.4.0      2019-10-04 [1]

@@

#>  iterators     1.0.12     2019-07-26 [1]

@@

#>  knitr         1.26       2019-11-12 [1]
#>  lattice       0.20-38    2018-11-04 [1]

@@

#>  Matrix      * 1.2-17     2019-03-22 [1]

@@

#>  mgcv          1.8-31     2019-11-09 [1]

@@

#>  nlme          3.1-142    2019-11-07 [1]

@@

#>  pkgbuild      1.0.6      2019-10-09 [1]

@@

#>  reprex        0.3.0      2019-05-16 [1]

@@

#>  rmarkdown     1.18       2019-11-27 [1]

@@

#>  rticles       0.12       2019-11-12 [1]

@@

#>  shape         1.4.4      2018-02-07 [1]

@@

#>  survival    * 3.1-7      2019-11-09 [1]
#>  testthat      2.3.1      2019-12-01 [1]

@@

#>  tidyverse   * 1.3.0      2019-11-21 [1]

@@

#>  xfun          0.11       2019-11-12 [1]

@@

#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          

@@

#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          

@@

\begin{CodeOutput}
#> Local:    master /Users/jislam/Desktop/Gits/cbpaper
#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
#> Head:     [71a5a9c] 2019-12-06: Lit-Review Edits
\end{CodeOutput}
\end{CodeChunk}

>>>>

To conduct our review on current methods available in survival analysis,
a scraper was coded to examine all packages in the survival task view on
CRAN \citeyearpar{survTaskView}. There was a total of 259 packages, many
of which are not relevant to the problems casebase addresses. To filter
these packages, a scraper was used to search for competing risks,
cumulative incidence functions, non- proportional hazards and
penalization. This reduced our set of 259 packages down to 60. From
here, we manually examined each package to determine which ones are
comparable to casebase, with either a different approach to the same
problems or future features we would like to add to casebase. In
particular, we excluded packages that were focused on a different set of
problems, such as frailty and multi-state models, as casebase is
currently focused on absolute-risk, competing events and single event
analysis.

These packages can belong to three main categories: non-proportional
hazards, regularization and competing risks. Besides these three main
categories, there are extra features of interest as well. Table
\ref{tab:surv-pkgs} summarizes the findings.

Many packages exist that implement semi-parametric Cox models. Of those,
\pkg{glmnet} \citeyearpar{regpathcox}???, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche} introduce penalization
in their model fitting. \pkg{CoxRidge} \citeyearpar{perperoglou},
\pkg{rstpm2} \citeyearpar{clements_liu} and \pkg{survival}
\citeyearpar{survival-package} include penalization, but they implement
the extended cox model permitting intuitive non proportional hazard
modeling while the previous four do not. In terms of penalization, the
benefits of \pkg{casebase} comes from its flexibility. Linear models
have a wide tool set, considerably larger than those made for Cox
regression. \pkg{casebase} permits many regularization tools developed
for linear regression to be repurposed within a survival context.

Parametric models permit non-proportional hazards and are implemented in
a handful of packages. Of interest are \pkg{CFC}
\citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu} and \pkg{survival}
\citeyearpar{survival-package}. \pkg{SmoothHazard} is limited to Weibull
distributions \citeyearpar{smoothHazard}, whereas \pkg{flexsurv} and
\pkg{survival} contain implementations for users to supply any
distribution they would like \citeyearpar{survival-package},
\citeyearpar{flexsurv}. \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets}
and \pkg{rstpm2} can use splines, permitting a flexible fit over time
\citeyearpar{flexsurv}, \citeyearpar{smoothHazard},
\citeyearpar{scheike2014estimating}, \citeyearpar{clements_liu}. Though
\pkg{casebase} is limited to three distributions at the moment, it also
uses splines allowing for distributions that may not be consistent over
survival time.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
\pkg{mets} and \pkg{survival} contain implementations for competing
risks \citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
\citeyearpar{scheike2014estimating}, \citeyearpar{survival-package}. The
differentiating factor between these packages and \pkg{casebase} is that
\pkg{casebase} handles competing risks through multiple logistic
regression. \pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can
handle interval censoring, whereas \pkg{casebase} currently cannot.
Interval censoring is a simple feature to implement and is a potential
improvement to add to \pkg{casebase}.

Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC},
\pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and
\pkg{survival} have implementations for cumulative incidence
\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche},
\citeyearpar{clements_liu}, \citeyearpar{survival-package}. Since
\pkg{casebase} is parametric, the baseline hazard is known and requires
integration to retrieve the cumulative incidence. This is similar to the
approaches of \pkg{CFC} and \pkg{flexsurv}
\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv}, while the
other packages use semi-parametric approaches to retrieve the baseline
after fitting \citeyearpar{scheike2014estimating},
\citeyearpar{gerds_blanche}. \pkg{rstpm2} and \pkg{survival} use both
for their respective parametric and semi-parametric model
implementations \citeyearpar{clements_liu},
\citeyearpar{survival-package}.

\begin{table}[ht]

@@

This report was generated on 2019-12-11 15:31:05 using the following
computational environment and dependencies:

@@

#> ─ Session info ──────────────────────────────────────────────────────────

@@

#>  tz       America/Winnipeg            

@@

#> ─ Packages ──────────────────────────────────────────────────────────────

@@

#>  callr         3.3.1      2019-07-18 [1]

@@

#>  codetools     0.2-16     2018-12-24 [4]

@@

@@

#>  devtools      2.1.0      2019-07-06 [1]

@@

@@

#>  foreach     * 1.4.4      2017-12-12 [1]

@@

#>  glmnet      * 2.0-18     2019-05-20 [1]

@@

#>  haven         2.1.1      2019-07-04 [1]

@@

#>  htmltools     0.3.6      2017-04-28 [1]

@@

#>  iterators     1.0.9      2017-12-12 [1]

@@

#>  knitr         1.23       2019-05-18 [1]
#>  lattice       0.20-38    2018-11-04 [4]

@@

#>  Matrix      * 1.2-18     2019-11-27 [4]

@@

#>  mgcv          1.8-24     2018-06-18 [1]

@@

#>  nlme          3.1-142    2019-11-07 [4]

@@

#>  pkgbuild      1.0.3      2019-03-20 [1]

@@

@@

#>  rmarkdown     1.16       2019-10-01 [1]

@@

#>  rticles       0.9.1      2019-07-22 [1]

@@

@@

#>  survival    * 2.44-1.1   2019-04-01 [4]
#>  testthat      2.2.0      2019-07-22 [1]

@@

#>  tidyverse   * 1.2.1      2017-11-14 [1]

@@

#>  xfun          0.8        2019-06-25 [1]

@@

#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          

@@

#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.2)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.2)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.5.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.2)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  Github (rstudio/rticles@8d56fc6)        
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.2)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.1)                          

@@

\begin{CodeOutput}
#> Local:    Jesse-Islam-master /home/mturgeon/Documents/git_repositories/cbpaper
#> Head:     [32c45a7] 2019-12-08: fix table captions
\end{CodeOutput}
\end{CodeChunk}
